---
title: 'Data wrangling for spatial analysis: R Workshop'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(raster)
library(leaflet)
library(tmap)
library(readr)
library(concaveman)
library(mgcv)
```

```{r}
setwd("~/Doctorat/Formation/Data_wrangling_for_spatial_analysis")
```
Loading data
```{r}
dat<-read_csv("data-for-course/copepods_raw.csv")
```
1.1 Shape data and maps ==============================================================================================
Initial visuals
```{r}
ggplot(dat)+
  aes(x=longitude,y=latitude,color=richness_raw)+
  geom_point()
```
```{r}
ggplot(dat)+
  aes(x = latitude,y=richness_raw)+
  geom_point()+
  stat_smooth()+
```
Introduction to maps
```{r}
sdat<-st_as_sf(dat,
               coords = c("longitude","latitude"),
               crs="+proj=longlat +datum=WGS84 +nodefs")
```
sf=simple features
sf point collection
```{r}
sdat
```
basic cartography
```{r}
plot(sdat["richness_raw"])
```
Thematic maps
```{r}
tm_shape(sdat)+
  tm_dots(col="richness_raw")
```
Customize previous plot
```{r}
tm1 <- tm_shape(sdat) +
  tm_dots(col = "richness_raw",
          palette="Blues",
          title="Species nb")
tm1
```
Save map
```{r}
tmap_save(tm1,
          filename="Richness_map.png",
          width=600,
          height=600)
```
Create a sf polygon collection
Download data from AODN
```{r}
shelfin<-st_read("data-for-course/data_raw/contour_5m.shp")
shelfin
range(shelfin$CONTOUR)
shelf200<-dplyr::filter(shelfin, CONTOUR == 199.9)
plot(shelf200["LENGTH"])
```
Create a polygon from lines
```{r}
ushelf200<-concaveman(shelf200)
plot(ushelf200)
st_is_valid(ushelf200)
sf::sf_use_s2(FALSE)
ushelf200 <- st_buffer(ushelf200, 0.0)
ushelf200 <- st_zm(ushelf200)
st_write(ushelf200, "data-for-course/spatial-data/aus_shelf.shp")
```

```{r}
aus<-st_read("data-for-course/spatial-data/AUS_2021_AUST_GDA2020.shp")
shelf <- st_read("data-for-course/spatial-data/aus_shelf.shp")
shelf
tm_shape(shelf)+
  tm_polygons()
```
ADD layers to a polygon
```{r}
tmap_mode("view")
tm_shape(shelf,bbox=sdat)+
  tm_polygons(col='blue')+
  tm_shape(aus$geometry)+
  tm_polygons()+
  tm_shape(sdat)+
  tm_dots()+
  tmap_style("natural")

tm_shape(shelf,bbox=sdat,)+
  tm_borders(col='blue')+
  tm_shape(aus$geometry)+
  tm_borders()+
  tm_shape(sdat)+
  tm_dots()+
  tm_basemap("Stamen.Watercolor")+
  tm_tiles("Stamen.TonerLabels")
```
Introduction to dplyr
Table joins with spatial data
```{r}
routes<-read_csv("data-for-course/Route-data.csv")
sdat_std<-inner_join(sdat,routes,by="route")
nrow(sdat)
nrow(sdat_std)
nrow(routes)
length(unique(routes$route))
```
Adding new variables
```{r}
sdat_std <-  mutate(sdat_std,
              richness = richness_raw/silk_area)
sdat_std$Lat <- st_coordinates(sdat_std)[,2]
ggplot(sdat_std)+
  aes(x=Lat,y=richness,color=richness)+
  geom_point()+
  stat_smooth()+
  theme_bw()
save(sdat_std,file =  "data-for-course/spatial-data/copepods_standardised.rda")
```

1.2 GIS and data analysis ============================================================================================
Changes Coordinates system
```{r}
shelf <- st_read("data-for-course/spatial-data/aus_shelf.shp")
shelf$shelf <- "Shelf"
st_crs(sdat_std)==st_crs(shelf)
shelf <- st_transform(shelf, crs = st_crs(sdat_std))
```
Intersecting points and polygons
```{r}
sdat_shelf <- st_join(sdat_std, shelf, join = st_intersects)
names(sdat_shelf)
unique(sdat_shelf$shelf)
sdat_shelf <- mutate(sdat_shelf,
                     shelf = if_else(is.na(shelf),
                                     "Offshore",
                                     shelf))
table(sdat_shelf$shelf)
```
Mapping data of two levels
```{r}
tm_shape(shelf,
         bbox = sdat_shelf)+
  tm_polygons(col="grey10")+
  tm_shape(sdat_shelf)+
  tm_dots(col="shelf",palette="RdBu")+
  tm_graticules()
```
Analysis of richness by continental shelf/offshore
```{r}
ggplot(sdat_shelf)+
  aes(x=Lat,y=richness,color=shelf)+
  geom_point(alpha=0.5,size=0.2)+
  stat_smooth()+
  theme_bw()
```
Introduction to raster data
```{r}
rsst <- raster('data-for-course/spatial-data/MeanAVHRRSST.grd')
plot(rsst)
```
Extracting temperatures at the sampling sites
```{r}
sdat_std$sst <- extract(rsst, sdat_std)
ggplot(sdat_std, aes(sst, richness)) +
  geom_point() +
  theme_minimal()
with(sdat_std, cor.test(sst, richness))
filter(sdat_std, is.na(sst))
sdat_sst <- filter(sdat_std, !is.na(sst))
```
Simple model of SST
```{r}
m1 <- gam(richness ~ s(sst, k=5), data = sdat_sst, family = 'poisson')
plot(m1)
```
Accounting for regions
```{r}
sdat_sst$Region <- factor(sdat_sst$region)
m1 <- gam(richness ~ s(sst, k=5, by = Region) + Region, data = sdat_sst, family = 'poisson')
sdat_sst$pred_m1 <- predict(m1, type = "response")
ggplot(sdat_sst) +
  aes(x = sst, y = richness, color = Region)+
  geom_point(size = 0.2, alpha = 0.3) +
  geom_line(aes(y = pred_m1), size = 1) +
  facet_grid(.~Region) +
  theme_bw()
```
Accounting for overdispersion
```{r}
deviance(m1)
m1$df.residual
m2 <- gam(richness ~ s(sst, by = Region) + Region, data = sdat_sst, family = mgcv::negbin(theta = 1.99))
deviance(m2)
m2$df.residual
```
Rootograms for checking model fit
```{r}
sdat_sst$pred_m2 <- predict(m2, type = "response")

ggplot(sdat_sst) +
  aes(x = sst, y = richness, color = Region)+
  geom_point(size = 0.2, alpha = 0.3) +
  geom_line(aes(y = pred_m2), size = 1) +
  facet_grid(.~Region) +
  theme_bw()
```

Spatial models of SST and richness
Using longitude as a covariate
```{r}
sdat_sst$x <- st_coordinates(sdat_sst)[,1]
m_int <- gam(richness ~s(sst, x), data = sdat_sst, family = mgcv::negbin(theta = 2.03))

m_int$df.residual
deviance(m_int)
plot(m_int)
```
Challenge topic: Spatial ‘clumping’ model for the West Coast
```{r}
sdat_sst$x <- st_coordinates(sdat_sst)[,1]
sdat_sst$y <- st_coordinates(sdat_sst)[,2]

sdat_west <- filter(sdat_sst, (x < 120) & (y > -40))
sdat_sst$group <- cut(sdat_sst$y, breaks = seq(-37, -16, by = 1.5))
m4 <- gamm(richness ~ s(sst), data = sdat_sst,
           random = list(group=~1),
          family = mgcv::negbin(theta = 3.8))
plot(m4$gam)
summary(m4$lme)
```
Challenge topic: Exploratory model with x and y as smoothers
```{r}
sdat_west2 <- st_transform(sdat_west, crs = 3577)
sdat_west2$x <- st_coordinates(sdat_west2)[,1]
sdat_west2$y <- st_coordinates(sdat_west2)[,2]

m5 <- gam(richness ~ s(sst) + s(x, y, bs = "gp"),
          data = sdat_west2,
          family = mgcv::negbin(theta = 3.8))
plot(m5)
```
Challenge topic: Modelling spatial autocorrelation
```{r}
m6 <- gamm(richness ~ s(sst),
                #This next step includes the spatial AC
                # with an exponential correlation structure
                correlation = corExp(form = ~x + y),
                data = sdat_west2, family =
                  mgcv::negbin(theta = 3.8))

plot(m6$gam)

summary(m6$lme)
```
Generating and mapping model predictions
```{r}
sdat_sst$richness_pred <- predict(m_int, type = "response")
tm_shape(sdat_sst) +
  tm_dots(col = "richness_pred") 
```

Generating predictions anywhere
```{r}
rsst2 <- aggregate(rsst, 2)
par(mfrow = c(1,2))
plot(rsst)
plot(rsst2)
```
```{r}
icell <- 1:ncell(rsst)
pred <- data.frame(sst = rsst2[icell],
                       cells = icell,
                       x = xFromCell(rsst2, icell),
                      y = yFromCell(rsst2, icell))

pred <- na.omit(pred)
head(pred)
```

```{r}
pred$richness_pred <- predict(m_int, newdata = pred, type = "response")
head(pred)
rpred <- raster(rsst2)
rpred[pred$cells] <- pred$richness_pred
tm_shape(aus, bbox = rpred) +
  tm_polygons(col = "white") +
  tm_shape(rpred) +
  tm_raster(palette = "RdPu",
            title= "Richness", alpha = 0.8, n=10) +
  tm_layout(bg.color = "grey20",
            legend.position = c("left", "top"),
            legend.text.color = "white",
            legend.title.color = "white")
```
```{r}
ggplot(sdat_sst) +
  aes(x = sst, y = richness_pred, color = x, alpha = 0.5) +
  geom_point() +
    theme_bw() +
  ylab("Richness (predicted)") +
  xlab(expression('SST ('*~degree*C*')'))
```
Create an interactive map
```{r}
print(object.size(sdat_sst), units = "Kb")
leaflet(sdat_sst) %>%
  addTiles() %>%
  addCircleMarkers(radius = 0.5)
```

```{r}
copedomain <- range(sdat_sst$richness)
oranges <- colorNumeric("YlOrRd", domain = copedomain)
leaflet(sdat_sst) %>%
  addProviderTiles("Esri.OceanBasemap") %>%
  addCircleMarkers( radius = 3,
                   color = 'grey80',
                   weight = 0.1,
                   fill = TRUE,
                   fillOpacity = 0.7, fillColor = ~oranges(richness)) %>%
  addLegend("topright", pal = oranges,
values = copedomain,
title = "Number of copepod species",
opacity = 1) 
```
Bonus material: Changing the map projection
```{r}
tmaptools::get_proj4("robin", output = "character")
robin <-  "+proj=robin +lon_0=100 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs +towgs84=0,0,0"
tm_shape(rsst, projection = robin) +
  tm_raster(palette = "-RdBu", title = "SST") +
  tm_shape(aus, projection = robin) +
  tm_polygons(col = "wheat") +
  tm_compass() 
```
Bonus material: calculating distances with sf
```{r}
tas_ext <- extent(140, 155, -45, -39)
stas <- st_crop(sdat_shelf, tas_ext)
```

Slow but precise way
```{r}
dist <- st_distance(stas, shelf)
```

Fast but less precise way
```{r}
tas_utm <- crs("+proj=utm +zone=55 +datum=WGS84 +units=m +no_defs")
stas2 <- st_transform(stas, crs = tas_utm)
shelf2 <- st_crop(shelf, tas_ext)
shelf2 <- st_transform(shelf2, crs = tas_utm)

dist2 <- st_distance(stas2, shelf2)
stas2$dist <- as.numeric(dist2)/1000
```
Plot samples by their distance to the shelf
```{r}
tm_shape(stas2) +
  tm_dots() +
  tm_shape(shelf2) +
  tm_polygons(col = "lightblue") +
  tm_shape(stas2) +
  tm_symbols(col = "dist", alpha = 0.5,
             title.col = "Distance from \n shelf (km)")
```
```{r}
tm_shape(shelf2) +
  tm_polygons(col = "lightblue") +
tm_shape(aus, projection = tas_utm) +
  tm_polygons() +
  tm_shape(stas2) +
  tm_symbols(col = "dist", alpha = 0.5,
             title.col = "Distance from \n shelf (km)")
```
```{r}
ggplot(stas2) +
  aes(x = dist, y = richness) +
  geom_point() +
  stat_smooth()
```

